---
title: "Study of Temporal and Spatial Generation Patterns of U.S. Power Plants"
author: "Xin Zhang"
fontsize: 12pt
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
  word_document: default
mainfont: Times New Roman
geometry: margin=2.54cm
subtitle: https://github.com/zhangxsaras/eGRID16
abstract: Power plant emissions contribute a great amount Green House Gas (GHG) to the atmosphere and GHG emissions account a lot for the global mean temperature rise. Power Plant emission amount is related to its generation and in general, higher generation indicates higher fuel use and higher emission. Studying power plants generation can help us understand emission pattern. To study the temporal and spatial pattern of power plants generation can help us predict future power generation and then we can give suggestions accordingly to reduce emission. From the time series analysis and spatial analysis results, it turned out there is an installation time trend for generator annual net generation, and there are spatial distribution patterns of power plants in U.S.. From 1900s to 2016, the generation increased with fluctuation and the trend was related to policy and world fuel market. We could not predict the future electricity generation since there was no clear trend after 1987. However, from the differences of spatial distribution patterns of Power Plants in U.S., we found that using renewable energy to replace traditional energy power plants can help a lot to reduce GHG emission in the future.
editor_options: 
  chunk_output_type: console
---

<Information in these brackets are used for annotating the RMarkdown file. They will not appear in the final version of the PDF document>

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

<Setup the global options for the R chunks in your document>

<Note: set up autoreferencing for figures and tables in your document>

```{r setup, include=FALSE}
# Set your working directory
getwd()

# Load your packages
library(tidyverse)
library(lubridate)
library(nlme)
library(lsmeans)
library(multcompView)
library(trend)
library(sf)
library(leaflet)
library(mapview)
library(RColorBrewer)
library(usmap)

# Set your ggplot theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```


# Research Question and Rationale

<Paragraph detailing the rationale for your analysis. What is the significant application and/or interest in this topic? Connect to environmental topic(s)/challenge(s).>

<Paragraph detailing your research question(s) and goals. What do you want to find out? Include a sentence (or a few) on the dataset you are using to answer this question - just enough to give your reader an idea of where you are going with the analysis.>

>Public awareness of global climate change’s impacts on the environment and the society has increased. Greenhouse gas (GHG) emissions account a lot for the global mean temperature rise. To achieve the goal of limiting global warming to 1.5°C, by 2030, global net human-caused emissions of carbon dioxide (CO2) would need to decrease by around 45 % from 2010 levels, and by 2050, CO2 emission would need to reach ‘net zero’ (IPCC, 2018). In United States, a large amount of GHG emissions come from the electric power sector. The electric power sector accounts for 38% of the total U.S. energy consumption in 2017 (US EIA, 2017). In the same year, the transportation end-use sector and other mobile combustion accounted for 1,732.02 MMT CO2Eq., which accounts 28% of the total GHG emission in the U.S., which is the second largest GHG emissions from all other sectors (the largest is the transportation sector) (US EPA, 2018). To study the temporal and spatial pattern of power plants generation can help us predict future power generation and then we can give suggestions accordingly to reduce emission. 

>Therefore, here I raise two research questions:
>(1) Is there an installation time trend for generator annual net generation? (New power generators tend to have higher or lower capacity overtime?) 
>(2) Is there a spatial distribution pattern of Power Plants in U.S.? (Number of power plants/ Total Electricity Generation/ Equivalent CO2 emission in each state ~ States)

>Here I am going to use the 2016 Emissions & Generation Resource Integrated Database (eGRID). It is a comprehensive source of data on the environmental characteristics of # almost all electric power generated in the United States. I am going to use the GEN16 tab (2016 Generators) and the PLNT 16 tab (2016 Plants). (Note: usually one power plant has several generators). The former gives information of all the generators currently in use in 2016 and the latter gives information of all the power plants in use in 2016. 

\newpage

# Dataset Information

<Information on how the dataset for this analysis were collected, the data contained in the dataset, and any important pieces of information that are relevant to your analyses. This section should contain much of same information as the README file for the dataset but formatted in a way that is more narrative.>

<Add a table that summarizes your data structure. This table can be made in markdown text or inserted as a `kable` function in an R chunk. If the latter, do not include the code used to generate your table.>

> The dataset was retrieved at:
> https://www.epa.gov/energy/emissions-generation-resource-integrated-database-egrid
> on 2019-03-25 21:10:11 EDT. It was originally an excel file and I saved it as two separate csv files of GEN16 an PLNT16. Since the dataset has two names in the first two rows (one full name and one abbreviate one), here I only list the columns I will use later, others can check the original dataset for references.

> Dataset Name | Information | Useful Column1 | Useful Column2 | Useful Column3 | Useful Column4 
> ------------- | ------------- | ------------- | ------------- | ------------- | -------------
> GEN16 | Generators Information | SEQGEN16: eGRID2016 Plant file sequence number | PSTATABB: Plant state abbreviation | GENNTAN: Generator annual net generation (MWh) | GENYRONL: Generator year on-line
> PLNT16 | Plants Information | SEQGEN16: eGRID2016 Plant file sequence number | PSTATABB: Plant state abbreviation | PLNGENAN: Plant annual net generation (MWh) | PLCO2EQA: Plant annual CO2 equivalent emissions (tons)

```{r, }


```


\newpage

# Exploratory Data Analysis and Wrangling

<Include R chunks for 5+ lines of summary code (display code and output), 3+ exploratory graphs (display graphs only), and any wrangling you do to your dataset(s).> 
```{r}
#Load dataset
GEN16 <- read.csv("./Data/Raw/egrid_GEN16.csv")
PLNT16 <- read.csv("./Data/Raw/egrid_PLNT16.csv")

#data wrangling
#This dataset has two names, one is the full name, and the first row is the abbreviation
#change column names to the abbr.
names(GEN16) <- lapply(GEN16[1, ], as.character)
names(PLNT16) <- lapply(PLNT16[1, ], as.character)

#filter out the first row and use abbr. as column names
GEN16 <- GEN16[2:26184,] 
PLNT16 <- PLNT16[2:9710,]


#numeric data has comma, convert factor to numberic
class (GEN16$GENNTAN)
GEN16$GENNTAN <-as.numeric(gsub(",", "", GEN16$GENNTAN))
PLNT16$PLNGENAN <-as.numeric(gsub(",", "", PLNT16$PLNGENAN))
PLNT16$PLCO2EQA <-as.numeric(gsub(",", "", PLNT16$PLCO2EQA))

#Year data as.Date
class (GEN16$GENYRONL)
GEN16$GENYRONL<-as.Date(GEN16$GENYRONL,format = "%Y")

#filter data by the sequence of time
GEN16 = GEN16[order(GEN16[,'GENYRONL']),]

#GEN16 - sum the totoal generation by year
GEN16sel <- GEN16 %>%
  select(SEQGEN16, PSTATABB, GENNTAN, GENYRONL) %>%
  filter(!is.na(GENNTAN)) %>%
  filter(GENNTAN>0) %>%
  group_by(GENYRONL)%>%
  summarise(GENSUM = sum(GENNTAN)) 

#PLNT16 - sum the totoal generation/CO2/plant numbers by state
PLNT16sel <- PLNT16 %>%
  select(SEQPLT16, PSTATABB, PLPRMFL, PLNGENAN, PLCO2EQA) %>%
    filter(!is.na(PLNGENAN)&!is.na(PLCO2EQA)) %>%
  filter(PLNGENAN>0)%>%
  group_by(PSTATABB)%>%
  summarise(PLNTGEN = sum(PLNGENAN),
            ECO2 = sum(PLCO2EQA),
            Count=n())

#summary code for GEN16
colnames(GEN16sel)
class(GEN16sel$GENSUM)
class(GEN16sel$GENNTAN)
summary(GEN16sel)
dim(GEN16sel)
head(GEN16sel)

#summary code for PLNT16
colnames(PLNT16sel)
class(PLNT16sel$PLNTGEN)
class(PLNT16sel$ECO2)
summary(PLNT16sel)
dim(PLNT16sel)
head(PLNT16sel)

#save new datasets
#write.csv(GEN16sel, file = "./Data/Processed/GEN16sel_Processed.csv",row.names=FALSE)
#write.csv(PLNT16sel, file = "./Data/Processed/PLNT16sel_Processed.csv",row.names=FALSE)
#data prep for shiny app
#PLNT16orisel <- PLNT16 %>%
#  select(SEQPLT16, PSTATABB, PLPRMFL, PLNGENAN, PLCO2EQA) %>%
#    filter(!is.na(PLNGENAN)&!is.na(PLCO2EQA)) %>%
#  filter(PLNGENAN>0)
#write.csv(PLNT16orisel, file = "./Data/Processed/PLNT16ori_Processed.csv",row.names=FALSE)
```


```{r, echo=FALSE, fig.cap= "Overview of Generator online year"}
ggplot(GEN16sel, aes(x = GENYRONL, y = GENSUM)) + 
  geom_point()+
  labs(x= "Year", y="Totoal Annual Net Generation (MWh)")
```

```{r, fig.width=16, fig.height=8, echo=FALSE, fig.cap= "Plant Numbers"}
ggplot(PLNT16sel) + 
  geom_bar(aes(x = PSTATABB , y = Count),stat = "identity")+
  labs(x= "State", y="Total Plant Numbers")
```

```{r, fig.width=16, fig.height=8, echo=FALSE, fig.cap= "Plant Generation"}
ggplot(PLNT16sel) + 
  geom_bar(aes(x = PSTATABB , y = PLNTGEN),stat = "identity")+
  labs(x= "State", y="Totoal Plant Net Generation (MWh)")
```

```{r, fig.width=16, fig.height=8, echo=FALSE, fig.cap= "Plant Equivalent CO2 Emissions"}
ggplot(PLNT16sel) + 
  geom_bar(aes(x = PSTATABB , y = ECO2),stat = "identity")+
  labs(x= "State", y="CO2 equivalent emissions (tons)")
```

<Include text sections to accompany these R chunks to explain the reasoning behind your workflow, and the rationale for your approach.>


\newpage

# Analysis
<Include R chunks for 3+ statistical tests (display code and output) and 3+ final visualization graphs (display graphs only).>

<Include text sections to accompany these R chunks to explain the reasoning behind your workflow, rationale for your approach, and the justification of meeting or failing to meet assumptions of tests.>

```{r}
#Q1 Time series analysis on GEN16
# Use GLM to see if there is a significant time trend
GENTest.fixed <- gls(data = GEN16sel,
                     GENSUM ~ GENYRONL, 
                      method = "REML")
summary(GENTest.fixed) # significnat trend t=10.23, p<0.005.
```

> According to GLM, time has a significant effect on the total annual net generation of generators (t=10.23, p<0.05).

```{r}
# Run a Mann-Kendall test
mk.test(GEN16sel$GENSUM)
# there is a trend over time according to this test (p<0.001), Z is positive, positive trend - time passes, value increases 

# Test for change point
pettitt.test(GEN16sel$GENSUM) #changing point at 58 - Year 1952
#GEN16sel[58,]

# Run separate Mann-Kendall for each change point
mk.test(GEN16sel$GENSUM[1:57])
# there is a trend over time according to this test (p<0.001), Z is positive, positive trend - time passes, value increasess 
mk.test(GEN16sel$GENSUM[58:123])
# there is a trend over time according to this test (p<0.05), Z is positive, positive trend - time passes, value increasess 

# Is there a second change point?
pettitt.test(GEN16sel$GENSUM[58:123]) #there is! 17 - Year 1969
#GEN16sel[75,]

# Run separate Mann-Kendall for each change point
mk.test(GEN16sel$GENSUM[58:74])
# there is a trend over time according to this test (p<0.05), Z is positive, positive trend - time passes, value increasess 
mk.test(GEN16sel$GENSUM[75:123])
# there is a trend over time according to this test (p<0.05), Z is negative, negative trend - time passes, value decreases 

# Is there a third change point?
pettitt.test(GEN16sel$GENSUM[75:123]) #there is! 1987
# GEN16sel[94,]
mk.test(GEN16sel$GENSUM[75:93]) 
# no trend!
mk.test(GEN16sel$GENSUM[94:123])
# no trend!
# no trend before and after the changing point

```
> According to the Mann-Kendall test, there is a trend over time for total annual net generation of generators (p<0.001), Z =11.311, which indicates a positive trend over time. Accroding to pettitt test. There are three changing points: Year 1952, Year 1969 and Year 1987 (p<0.05).

```{r, fig.cap="Generators Time Series Analysis", echo=FALSE}
ggplot(GEN16sel, aes(x = GENYRONL, y = GENSUM)) + 
  geom_point() +
  labs(x= "Year", y="Totoal Annual Net Generation (MWh)")+
  geom_vline(xintercept = as.Date("1952-01-01"), color="#253494", lty = 2)+ 
  geom_vline(xintercept = as.Date("1969-01-01"), color="#253494", lty = 2)+
  geom_vline(xintercept = as.Date("1987-01-01"), color="#253494", lty = 2)+
  geom_text(x = as.Date("1952-01-01"), y = 80000000, label = "(1952; 6,177,035 MWh)", hjust = 1)+
  geom_text(x = as.Date("1969-01-01"), y = 120000000, label = "(1969; 59,997,590 MWh)", hjust = 1)+
  geom_text(x = as.Date("1987-01-01"), y = 200000000, label = "(1978; 70,272,900 MWh)", hjust = 1) 
```

> As is shown in the figure, before the first changing point 1952, the annual net generation of genrators in U.S. incrased very slowly each year and after 1952, the speed of generation change became faster. Annual net generation kept growing until 1969, this is the second change point. After 1969, there was a negative trend of annual net generation. The third changing point is Year 1987, and there was no clear pattern in 1969-1987 or after 1987.

```{r}
#Q2 spatial distribution analysis on PLNT16
#data wrangling
#add fips number to PLNT16sel data
library(usmap)
state_map <- us_map(regions = "states")
PLNT_State<- merge(state_map, PLNT16sel, by.x = "abbr", by.y = "PSTATABB")%>%
   select(abbr, fips, full, PLNTGEN, ECO2, Count)
PLNT_State = PLNT_State[!duplicated(PLNT_State$abbr),]

#join PLNT16 data to counties map data
states_sf<- st_read('./Data/RAW/States.shp') 
st_crs(states_sf)

PLNT_State_merge <- merge(states_sf, PLNT_State, by.x = "STATEFP", by.y = "fips")
#mapview(PLNT_State_merge)
```

```{r, fig.cap="Map of Plant Numbers", echo=FALSE}
#Plot plant numbers
#mapview(PLNT_State_merge['Count'], layer.name = "Total Plant Numbers")
plot_usmap(data = PLNT_State, values = "Count", lines = "red") + 
    scale_fill_gradient("Total Plant Numbers",low='grey95',high='#132B43')+
  theme(legend.position = "right")
```

```{r, fig.cap="Map of Plant Generations", echo=FALSE}
#Plot PLNT Generation
#mapview(PLNT_State_merge['PLNTGEN'], layer.name = "Total Plant annual net generation (MWh)")
#ggplot() + 
#  geom_sf(data = PLNT_State_merge, aes(fill=PLNTGEN))+
#  scale_fill_gradient("Total Plant annual net generation (MWh)",low='white',high='darkblue')+ 
#  theme(legend.position = "right")
plot_usmap(data = PLNT_State, values = "PLNTGEN", lines = "red") + 
    scale_fill_gradient("Total Plant annual net generation (MWh)",low='grey95',high='#132B43')+
  theme(legend.position = "right")
```

```{r, fig.cap="Map of Equivalent CO2 Emissions", echo=FALSE}
#Plot equivalent CO2
#mapview(PLNT_State_merge['ECO2'], layer.name = "Total Plant annual CO2 equivalent emissions (tons)")
plot_usmap(data = PLNT_State, values = "ECO2", lines = "red") + 
    scale_fill_gradient("Annual CO2 equivalent emissions (tons)",low='grey95',high='#132B43')+
  theme(legend.position = "right")
```

> Since our research scale is U.S., it also includes Alaska and this made it very difficult to map it using ggplot (AL will be far away from other sates and the whole map will be very small to show). Therefore, here I use the plot_usmap funciton instead.As the maps show (Figure 6, 7, 8), there are spatial distribution patterns. For total plant numbers, California ranks first but in general,there are more power plants in east cosat states than in west coast. Texas is another exception in the southern part that has relatively larger numbers of power plants.  For plant annual net generation, in general, east coast states also have relatively higher net generation than west coast states, and Texas ranks first. Accordingly, Texas also has highest annual CO2 equivalent emissions, and east coat states have higher emissions than west coast states.

\newpage

# Summary and Conclusions
<Summarize your major findings from your analyses. What conclusions do you draw from your findings? Make sure to apply this to a broader application for the research question you have answered.>

> From the analysis above, we can see that time has a significant effect on the total annual net generation of generators. There is a trend over time for total annual net generation of generators that new power generators tend to have higher or net generation overtime.

> There are also three changing points of the whole period: 1952, 1969, 1987. After 1952, there is a huge increase in power generation each year, this might be related to the technology thrive at the end of 1940s. The first hydraulic fracturing treatment was pumped on a gas well operated by Pan American Petroleum Corp in the Hugoton field in 1947 and it was the beginning of period of rapid electric industry growth (Oilscams.org, Access: 2019-4-14). However, at the end of 1960s, energy crisis caused the reduction of electricity generation and this energy crisis peaked at 1973 (Energy Crisis (1970s), Access: 2019-4-14). This accord with our second changing point here at 1969 and after 1969, there was a negative trend of annual net generation. 

> In the late 1970s, the government published more regulations on the electricity generation industry such as the National Energy Act in 1978 to exert more control on the electricity industry. In 1990s, the government published Congress passes Bush's Energy Policy Act (EPACT) to deregulate the electricity industry (Ballotpedia, Access: 2019-4-14). The back and force between strict and loose policies led to the fluctuations of the generator generations, and this might be a cause why there is a changing point of 1987 according to the pettitt test but there were no trend before and after this year according to the Mann-Kendall test.

> From the maps, we can see that there was spatial distribution pattern for power plants generation in 2016. In general, eastern states had higher power plant numbers, electricity net generation, and equivalent CO2 emissions than western states. California ranked first for total power plants nubmers but its anual net generation and equivalent CO2 emission were both much less than Texas.The main reason for this was the type of power plants. In Texas, there are more traditonal power plants using coal and natural gas as fuel, but in California, there are more renewable energy power plants. Compared to traditional power plants (600 MW, 0.75), renewable power plants have lower nameplate capacity and capacity factor (1000MW, 0.1). Therefore, even though California had more power plants, it had lower electricity generation. Also, compared to traditional power plants, renwable power plants use clean energy and barly have any emissions, so California's equivalent CO2 emissions were also much lower than Texas.

> To sum up, there is an installation time trend for generator annual net generation. From 1900s to 2016, the generation increased with fluctuation and the trend was related to policy and world fuel market. It was very difficult to predict the futhure electricity generation since there was no clear trend after 1987. However, from the differences of spatial distribution patterns of Power Plants in U.S., we can give suggesitons of using renewable energy to replace traditional energy power plants. This can help a lot to reduce GHG emission in the futher. 


