---
title: "eGRID Data Analysis"
author: "Xin Zhang"
fontsize: 12pt
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
  word_document: default
mainfont: Times New Roman
geometry: margin=2.54cm
subtitle: https://github.com/zhangxsaras/eGRID16
abstract: Experimental overview. This section should be no longer than 250 words.
editor_options: 
  chunk_output_type: console
---

<Information in these brackets are used for annotating the RMarkdown file. They will not appear in the final version of the PDF document>

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

<Setup the global options for the R chunks in your document>

<Note: set up autoreferencing for figures and tables in your document>

```{r setup, include=FALSE}
# Set your working directory
getwd()

# Load your packages
library(tidyverse)
library(lubridate)
library(nlme)
library(lsmeans)
library(multcompView)
library(trend)
library(sf)
library(leaflet)
library(mapview)
library(RColorBrewer)
library(usmap)

# Set your ggplot theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```


# Research Question and Rationale

<Paragraph detailing the rationale for your analysis. What is the significant application and/or interest in this topic? Connect to environmental topic(s)/challenge(s).>

<Paragraph detailing your research question(s) and goals. What do you want to find out? Include a sentence (or a few) on the dataset you are using to answer this question - just enough to give your reader an idea of where you are going with the analysis.>

\newpage

# Dataset Information

<Information on how the dataset for this analysis were collected, the data contained in the dataset, and any important pieces of information that are relevant to your analyses. This section should contain much of same information as the README file for the dataset but formatted in a way that is more narrative.>

<Add a table that summarizes your data structure. This table can be made in markdown text or inserted as a `kable` function in an R chunk. If the latter, do not include the code used to generate your table.>

\newpage

# Exploratory Data Analysis and Wrangling

<Include R chunks for 5+ lines of summary code (display code and output), 3+ exploratory graphs (display graphs only), and any wrangling you do to your dataset(s).> 
```{r}
#Load dataset
GEN16 <- read.csv("./Data/Raw/egrid_GEN16.csv")
PLNT16 <- read.csv("./Data/Raw/egrid_PLNT16.csv")

#data wrangling
#This dataset has two names, one is the full name, and the first row is the abbreviation
#change column names to the abbr.
names(GEN16) <- lapply(GEN16[1, ], as.character)
names(PLNT16) <- lapply(PLNT16[1, ], as.character)

#filter out the first row and use abbr. as column names
GEN16 <- GEN16[2:26184,] 
PLNT16 <- PLNT16[2:9710,]


#numeric data has comma, convert factor to numberic
class (GEN16$GENNTAN)
GEN16$GENNTAN <-as.numeric(gsub(",", "", GEN16$GENNTAN))
PLNT16$PLNGENAN <-as.numeric(gsub(",", "", PLNT16$PLNGENAN))
PLNT16$PLCO2EQA <-as.numeric(gsub(",", "", PLNT16$PLCO2EQA))

#Year data as.Date
class (GEN16$GENYRONL)
GEN16$GENYRONL<-as.Date(GEN16$GENYRONL,format = "%Y")

#filter data by the sequence of time
GEN16 = GEN16[order(GEN16[,'GENYRONL']),]

#GEN16 - sum the totoal generation by year
GEN16sel <- GEN16 %>%
  select(SEQGEN16, PSTATABB, GENNTAN, GENYRONL) %>%
  filter(!is.na(GENNTAN)) %>%
  filter(GENNTAN>0) %>%
  group_by(GENYRONL)%>%
  summarise(GENSUM = sum(GENNTAN)) 

#GEN16 - sum the totoal generation/CO2/plant numbers by state
PLNT16sel <- PLNT16 %>%
  select(SEQPLT16, PSTATABB, PLPRMFL, PLNGENAN, PLCO2EQA) %>%
    filter(!is.na(PLNGENAN)&!is.na(PLCO2EQA)) %>%
  filter(PLNGENAN>0)%>%
  group_by(PSTATABB)%>%
  summarise(PLNTGEN = sum(PLNGENAN),
            ECO2 = sum(PLCO2EQA),
            Count=n())

#summary code for GEN16
colnames(GEN16sel)
class(GEN16sel$GENSUM)
class(GEN16sel$GENNTAN)
summary(GEN16sel)
dim(GEN16sel)
head(GEN16sel)

#summary code for PLNT16
colnames(PLNT16sel)
class(PLNT16sel$PLNTGEN)
class(PLNT16sel$ECO2)
summary(PLNT16sel)
dim(PLNT16sel)
head(PLNT16sel)

#save new datasets
#write.csv(GEN16sel, file = "./Data/Processed/GEN16sel_Processed.csv",row.names=FALSE)
#write.csv(PLNT16sel, file = "./Data/Processed/PLNT16sel_Processed.csv",row.names=FALSE)
#data prep for shiny app
#PLNT16orisel <- PLNT16 %>%
#  select(SEQPLT16, PSTATABB, PLPRMFL, PLNGENAN, PLCO2EQA) %>%
#    filter(!is.na(PLNGENAN)&!is.na(PLCO2EQA)) %>%
#  filter(PLNGENAN>0)
#write.csv(PLNT16orisel, file = "./Data/Processed/PLNT16ori_Processed.csv",row.names=FALSE)
```


```{r, echo=FALSE}
ggplot(GEN16sel, aes(x = GENYRONL, y = GENSUM)) + 
  geom_point()+
  labs(x= "Year", y="Totoal Annual Net Generation (MWh)")
```

```{r, fig.width=16, fig.height=8, echo=FALSE}
ggplot(PLNT16sel) + 
  geom_bar(aes(x = PSTATABB , y = Count),stat = "identity")+
  labs(x= "State", y="Total Plant Numbers")

ggplot(PLNT16sel) + 
  geom_bar(aes(x = PSTATABB , y = PLNTGEN),stat = "identity")+
  labs(x= "State", y="Totoal Plant Net Generation (MWh)")

ggplot(PLNT16sel) + 
  geom_bar(aes(x = PSTATABB , y = ECO2),stat = "identity")+
  labs(x= "State", y="CO2 equivalent emissions (tons)")
```

<Include text sections to accompany these R chunks to explain the reasoning behind your workflow, and the rationale for your approach.>


\newpage

# Analysis
<Include R chunks for 3+ statistical tests (display code and output) and 3+ final visualization graphs (display graphs only).>

<Include text sections to accompany these R chunks to explain the reasoning behind your workflow, rationale for your approach, and the justification of meeting or failing to meet assumptions of tests.>

```{r}
#Q1 Time series analysis on GEN16
# Use GLM to see if there is a significant time trend
GENTest.fixed <- gls(data = GEN16sel,
                     GENSUM ~ GENYRONL, 
                      method = "REML")
summary(GENTest.fixed) # significnat trend t=10.23, p<0.005.
```

> According to GLM, time has a significant effect on the total annual net generation of generators (t=10.23, p<0.05).

```{r}
# Run a Mann-Kendall test
mk.test(GEN16sel$GENSUM)
# there is a trend over time according to this test (p<0.001), Z is positive, positive trend - time passes, value increases 

# Test for change point
pettitt.test(GEN16sel$GENSUM) #changing point at 58 - Year 1952
#GEN16sel[58,]

# Run separate Mann-Kendall for each change point
mk.test(GEN16sel$GENSUM[1:57])
# there is a trend over time according to this test (p<0.001), Z is positive, positive trend - time passes, value increasess 
mk.test(GEN16sel$GENSUM[58:123])
# there is a trend over time according to this test (p<0.05), Z is positive, positive trend - time passes, value increasess 

# Is there a second change point?
pettitt.test(GEN16sel$GENSUM[58:123]) #there is! 17 - Year 1969
#GEN16sel[75,]

# Run separate Mann-Kendall for each change point
mk.test(GEN16sel$GENSUM[58:74])
# there is a trend over time according to this test (p<0.05), Z is positive, positive trend - time passes, value increasess 
mk.test(GEN16sel$GENSUM[75:123])
# there is a trend over time according to this test (p<0.05), Z is negative, negative trend - time passes, value decreases 

# Is there a third change point?
pettitt.test(GEN16sel$GENSUM[75:123]) #there is! 1987
# GEN16sel[94,]
mk.test(GEN16sel$GENSUM[75:93]) 
# no trend!
mk.test(GEN16sel$GENSUM[94:123])
# no trend!
# no trend before and after the changing point

```
> According to the Mann-Kendall test, there is a trend over time for total annual net generation of generators (p<0.001), Z =11.311, which indicates a positive trend over time. Accroding to pettitt test. There are three changing points: Year 1952, Year 1969 and Year 1987 (p<0.05).

```{r}
ggplot(GEN16sel, aes(x = GENYRONL, y = GENSUM)) + 
  geom_point() +
  labs(x= "Year", y="Totoal Annual Net Generation (MWh)")+
  geom_vline(xintercept = as.Date("1952-01-01"), color="#253494", lty = 2)+ 
  geom_vline(xintercept = as.Date("1969-01-01"), color="#253494", lty = 2)+
  geom_vline(xintercept = as.Date("1987-01-01"), color="#253494", lty = 2)+
  geom_text(x = as.Date("1952-01-01"), y = 80000000, label = "(1952; 6,177,035 MWh)", hjust = 1)+
  geom_text(x = as.Date("1969-01-01"), y = 120000000, label = "(1969; 59,997,590 MWh)", hjust = 1)+
  geom_text(x = as.Date("1987-01-01"), y = 200000000, label = "(1978; 70,272,900 MWh)", hjust = 1) 
```

> As is shown in the figure, before the first changing point 1952, the annual net generation of genrators in U.S. incrased very slowly each year and after 1952, the speed of generation change became faster. Annual net generation kept growing until 1969, this is the second change point. After 1969, there was a negative trend of annual net generation. The third changing point is Year 1987, and there was no clear pattern in 1969-1987 or after 1987.

```{r}
#Q2 spatial distribution analysis on PLNT16
#data wrangling
#add fips number to PLNT16sel data
library(usmap)
state_map <- us_map(regions = "states")
PLNT_State<- merge(state_map, PLNT16sel, by.x = "abbr", by.y = "PSTATABB")%>%
   select(abbr, fips, full, PLNTGEN, ECO2, Count)
PLNT_State = PLNT_State[!duplicated(PLNT_State$abbr),]

#join PLNT16 data to counties map data
states_sf<- st_read('./Data/RAW/States.shp') 
st_crs(states_sf)

PLNT_State_merge <- merge(states_sf, PLNT_State, by.x = "STATEFP", by.y = "fips")
#mapview(PLNT_State_merge)

#Plot plant numbers
#mapview(PLNT_State_merge['Count'], layer.name = "Total Plant Numbers")
plot_usmap(data = PLNT_State, values = "Count", lines = "red") + 
    scale_fill_gradient("Total Plant Numbers",low='grey95',high='#132B43')+
  theme(legend.position = "right")

#Plot PLNT Generation
#mapview(PLNT_State_merge['PLNTGEN'], layer.name = "Total Plant annual net generation (MWh)")
#ggplot() + 
#  geom_sf(data = PLNT_State_merge, aes(fill=PLNTGEN))+
#  scale_fill_gradient("Total Plant annual net generation (MWh)",low='white',high='darkblue')+ 
#  theme(legend.position = "right")
plot_usmap(data = PLNT_State, values = "PLNTGEN", lines = "red") + 
    scale_fill_gradient("Total Plant annual net generation (MWh)",low='grey95',high='#132B43')+
  theme(legend.position = "right")

#Plot equivalent CO2
#mapview(PLNT_State_merge['ECO2'], layer.name = "Total Plant annual CO2 equivalent emissions (tons)")
plot_usmap(data = PLNT_State, values = "ECO2", lines = "red") + 
    scale_fill_gradient("Annual CO2 equivalent emissions (tons)",low='grey95',high='#132B43')+
  theme(legend.position = "right")
```

> Since our research scale is U.S., it also includes Alaska and this made it very difficult to map it using ggplot (AL will be far away from other sates and the whole map will be very small to show). Therefore, here I use the plot_usmap funciton instead.As the maps show, there are spatial distribution patterns. For total plant numbers, California ranks first but in general,there are more power plants in east cosat states than in west coast. Texas is another exception in the southern part that has relatively larger numbers of power plants.  For plant annual net generation, in general, east coast states also have relatively higher net generation than west coast states, and Texas ranks first. Accordingly, Texas also has highest annual CO2 equivalent emissions, and east coat states have higher emissions than west coast states.

\newpage

# Summary and Conclusions
<Summarize your major findings from your analyses. What conclusions do you draw from your findings? Make sure to apply this to a broader application for the research question you have answered.>
> From the analysis above, we can see that time has a significant effect on the total annual net generation of generators. There is a trend over time for total annual net generation of generators that new power generators tend to have higher or net generation overtime.
> There are also three changing points of the whole period: 1952, 1969, 1987. After 1952, there is a huge incrase in power generation each year, this might be related to the technology thirve at the end of 1940s. The first hydraulic fracturing treatment was pumped on a gas well operated by Pan American Petroleum Corp in the Hugoton field in 1947 and it was the beginning of period of rapid electric industry growth (Oilscams.org, Access: 2019-4-14). However, at the end of 1960s, energy crisis caused the reduction of electricity generation and this energ crsis peaked at 1973 (Energy Crisis (1970s), Access: 2019-4-14). This accord with our second changing point here at 1969 and after 1969, there was a negative trend of annual net generation. 
> In the late 1970s, the goverment published more regulations on the electricity generation industry such as the National Energy Act in 1978 to exert more control on the electricity indrustry. In 1990s, the government published Congress passes Bush's Energy Policy Act (EPACT) to deragulate the electricity indurstry (Ballotpedia, Access: 2019-4-14). The back and force between strict and loose policies led to the fluctuations of the generator generations, and this might be a cuase why there is a changing point of 1987 according to the pettitt test but there were no trend before and after this year according to the Mann-Kendall test.



